---
title: "VLK Fraud Detection — Local Outlier Factor"
author: "Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# 1. Load Libraries

```{r libraries}
# Install any missing packages before running:
# install.packages(c("tidyverse", "dbscan", "scales", "ggcorrplot", "uwot", "patchwork"))

library(tidyverse)    # data wrangling + ggplot2
library(dbscan)       # Local Outlier Factor (lof)
library(scales)       # axis formatting
library(ggcorrplot)   # correlation heatmap
library(uwot)         # UMAP dimensionality reduction
library(patchwork)    # combine ggplots
```

---

# 2. Load Data

```{r load-data}
df_raw <- read_csv("vlk_fraud 1.csv", show_col_types = FALSE)

cat("Rows:   ", nrow(df_raw), "\n")
cat("Cols:   ", ncol(df_raw), "\n")
cat("NAs:    ", sum(is.na(df_raw)), "\n")

head(df_raw)
```

```{r column-types}
glimpse(df_raw)
```

## 2.1 Filter Low-Value Transactions

We remove any transactions with an Amount of €0.99 or below. These sub-euro transactions are unlikely to represent meaningful fraudulent activity and add noise to the model.

```{r filter-low-value}
n_before <- nrow(df_raw)

df_raw <- df_raw %>%
  filter(Amount > 0.99)

n_after  <- nrow(df_raw)

cat("Removed:", n_before - n_after, "transactions (Amount ≤ €0.99)\n")
cat("Remaining:", n_after, "transactions\n")
```

---

# 3. Exploratory Data Analysis

## 3.1 Transaction Amount Distribution

```{r amount-dist}
ggplot(df_raw, aes(x = Amount)) +
  geom_histogram(bins = 100, fill = "#2c7bb6", color = "white", alpha = 0.85) +
  scale_x_log10(labels = dollar_format(prefix = "€")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Transaction Amount (log scale)",
    x     = "Amount (€, log scale)",
    y     = "Count"
  ) +
  theme_minimal(base_size = 13)

ggplot(df_raw, aes(x = Amount)) +
  geom_histogram(bins = 100, fill = "#2c7bb6", color = "white", alpha = 0.85) +
  scale_x_continuous(labels = dollar_format(prefix = "€")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Transaction Amount",
    x     = "Amount (€)",
    y     = "Count"
  ) +
  theme_minimal(base_size = 13)
```

```{r amount-summary}
summary(df_raw$Amount)
```

## 3.2 Feature Distributions (X01–X28)

Plotting the median and IQR for each feature to spot any that are extreme.

```{r feature-summary}
feature_cols <- paste0("X", sprintf("%02d", 1:28))

df_raw %>%
  select(all_of(feature_cols)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  group_by(feature) %>%
  summarise(
    mean   = mean(value),
    sd     = sd(value),
    min    = min(value),
    max    = max(value),
    .groups = "drop"
  ) %>%
  arrange(feature)
```

```{r feature-boxplots}
df_raw %>%
  select(all_of(feature_cols)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = feature, y = value)) +
  geom_boxplot(fill = "#abd9e9", outlier.size = 0.3, outlier.alpha = 0.3) +
  coord_flip() +
  labs(
    title = "Feature Distributions (X01–X28)",
    x     = NULL,
    y     = "Value"
  ) +
  theme_minimal(base_size = 11)
```

## 3.3 Correlation Heatmap

Since the features are already PCA-transformed, correlations should be low — but worth confirming.

```{r correlation}
cor_mat <- df_raw %>%
  select(all_of(feature_cols)) %>%
  cor()

ggcorrplot(cor_mat,
           type    = "lower",
           lab     = FALSE,
           colors  = c("#d7191c", "white", "#2c7bb6"),
           title   = "Feature Correlation Matrix") +
  theme(axis.text = element_text(size = 7))
```

---

# 4. Feature Engineering

The 28 features are already standardized (output of PCA). We only need to normalize `Amount` to the same scale before running LOF.

```{r feature-engineering}
df_model <- df_raw %>%
  mutate(Amount_scaled = as.numeric(scale(Amount))) %>%
  select(all_of(feature_cols), Amount_scaled)

cat("Feature matrix shape:", nrow(df_model), "x", ncol(df_model), "\n")
```

---

# 5. Local Outlier Factor (LOF)

## 5.1 What is LOF?

LOF compares the local density of a point to the densities of its neighbours. A score near 1 means the point is in a region of similar density (normal). A score **>> 1** means the point is much less dense than its neighbours — an outlier/anomaly.

## 5.2 Choosing k (number of neighbours)

A common heuristic for large datasets is k = 20. We'll also test k = 10 and k = 50 to assess sensitivity.

```{r lof-run}
set.seed(42)
feature_matrix <- as.matrix(df_model)

cat("Running LOF with k = 20 ...\n")
lof_k20 <- lof(feature_matrix, minPts = 20)

cat("Running LOF with k = 10 ...\n")
lof_k10 <- lof(feature_matrix, minPts = 10)

cat("Running LOF with k = 50 ...\n")
lof_k50 <- lof(feature_matrix, minPts = 50)

cat("Done.\n")
```

## 5.3 Score Distribution

```{r lof-score-dist}
score_df <- tibble(
  Transaction_ID = df_raw$Transaction_ID,
  Amount         = df_raw$Amount,
  lof_k10        = lof_k10,
  lof_k20        = lof_k20,
  lof_k50        = lof_k50
)

score_df %>%
  pivot_longer(starts_with("lof"), names_to = "k", values_to = "score") %>%
  ggplot(aes(x = score, fill = k)) +
  geom_histogram(bins = 100, alpha = 0.7, position = "identity") +
  scale_x_log10() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("#d7191c", "#2c7bb6", "#1a9641"),
                    labels = c("k=10", "k=20", "k=50")) +
  labs(
    title = "LOF Score Distribution (log scale)",
    x     = "LOF Score (log scale)",
    y     = "Count",
    fill  = NULL
  ) +
  theme_minimal(base_size = 13)
```

## 5.4 Score Correlation Across k Values

```{r lof-correlation}
score_df %>%
  select(starts_with("lof")) %>%
  cor() %>%
  round(4)
```

---

# 6. Ensemble Score & Ranking

We combine scores from k=10, 20, 50 using **rank averaging** — this is more robust than averaging raw scores directly (which can be dominated by extreme values).

```{r ensemble}
score_df <- score_df %>%
  mutate(
    rank_k10 = rank(lof_k10),
    rank_k20 = rank(lof_k20),
    rank_k50 = rank(lof_k50),
    ensemble_rank_score = (rank_k10 + rank_k20 + rank_k50) / 3
  ) %>%
  arrange(desc(ensemble_rank_score))

# Final ranked output
ranked_df <- score_df %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Transaction_ID, Amount, lof_k20, ensemble_rank_score)

head(ranked_df, 20)
```

---

# 7. Top 100 Suspicious Transactions

```{r top-100}
top100 <- ranked_df %>%
  slice_head(n = 100)

cat("Top 100 transactions selected.\n")
cat("LOF k=20 score range:", round(min(top100$lof_k20), 2), "–", round(max(top100$lof_k20), 2), "\n")
cat("Amount range: €", round(min(top100$Amount), 2), "–", round(max(top100$Amount), 2), "\n")
```

```{r top100-amount-dist}
ggplot(top100, aes(x = Amount)) +
  geom_histogram(bins = 30, fill = "#d7191c", color = "white", alpha = 0.85) +
  scale_x_continuous(labels = dollar_format(prefix = "€")) +
  labs(
    title = "Amount Distribution — Top 100 Flagged Transactions",
    x     = "Amount (€)",
    y     = "Count"
  ) +
  theme_minimal(base_size = 13)
```

```{r top100-score-dist}
ggplot(top100, aes(x = Rank, y = lof_k20)) +
  geom_col(fill = "#d7191c", alpha = 0.8) +
  labs(
    title = "LOF Scores for Top 100 Flagged Transactions",
    x     = "Rank",
    y     = "LOF Score (k=20)"
  ) +
  theme_minimal(base_size = 13)
```

---

# 8. Visualisation — UMAP Projection

We project all transactions to 2D with UMAP, then highlight the top 100 flagged transactions.

> **Note:** UMAP on 284k rows takes a few minutes. Set `n_threads` higher if your machine supports it.

```{r umap, cache=TRUE}
set.seed(42)
cat("Running UMAP (this may take a few minutes)...\n")

umap_result <- umap(
  feature_matrix,
  n_neighbors   = 15,
  min_dist      = 0.1,
  n_threads     = 4,
  verbose       = FALSE
)

umap_df <- tibble(
  Transaction_ID = df_raw$Transaction_ID,
  UMAP1          = umap_result[, 1],
  UMAP2          = umap_result[, 2]
) %>%
  left_join(score_df %>% select(Transaction_ID, ensemble_rank_score), by = "Transaction_ID") %>%
  mutate(
    flagged = Transaction_ID %in% top100$Transaction_ID,
    group   = if_else(flagged, "Top 100 Flagged", "Normal")
  )

cat("UMAP done.\n")
```

```{r umap-plot}
# Plot normal transactions first, flagged on top
p_umap <- ggplot() +
  geom_point(
    data  = filter(umap_df, !flagged),
    aes(x = UMAP1, y = UMAP2),
    color = "#aec6cf", size = 0.2, alpha = 0.3
  ) +
  geom_point(
    data  = filter(umap_df, flagged),
    aes(x = UMAP1, y = UMAP2),
    color = "#d7191c", size = 2.5, alpha = 0.9
  ) +
  labs(
    title    = "UMAP Projection — All Transactions",
    subtitle = "Red = Top 100 Flagged by LOF | Grey = Normal",
    x        = "UMAP Dimension 1",
    y        = "UMAP Dimension 2"
  ) +
  theme_minimal(base_size = 13)

p_umap
```

## 8.1 Zoom: LOF Score Coloured

```{r umap-score-colour}
# Sample 10,000 normal transactions to keep the plot readable
set.seed(42)
normal_sample <- umap_df %>% filter(!flagged) %>% slice_sample(n = 10000)

ggplot() +
  geom_point(
    data = normal_sample,
    aes(x = UMAP1, y = UMAP2, color = ensemble_rank_score),
    size = 0.5, alpha = 0.5
  ) +
  geom_point(
    data  = filter(umap_df, flagged),
    aes(x = UMAP1, y = UMAP2),
    color = "#d7191c", size = 2.5
  ) +
  scale_color_gradient(low = "#abd9e9", high = "#313695") +
  labs(
    title    = "UMAP — Anomaly Score Intensity",
    subtitle = "Red points = Top 100 | Colour = ensemble LOF rank score",
    color    = "Ensemble\nRank Score",
    x        = "UMAP Dimension 1",
    y        = "UMAP Dimension 2"
  ) +
  theme_minimal(base_size = 13)
```

---

# 9. Feature Analysis of Top 100

Compare the mean feature values of the top 100 vs. the rest of the dataset to understand *what makes them anomalous*.

```{r feature-comparison}
flagged_ids <- top100$Transaction_ID

comparison_df <- df_raw %>%
  mutate(group = if_else(Transaction_ID %in% flagged_ids, "Flagged Top 100", "Normal")) %>%
  select(group, all_of(feature_cols), Amount) %>%
  pivot_longer(-group, names_to = "feature", values_to = "value") %>%
  group_by(group, feature) %>%
  summarise(mean_val = mean(value), .groups = "drop") %>%
  pivot_wider(names_from = group, values_from = mean_val) %>%
  mutate(difference = `Flagged Top 100` - Normal) %>%
  arrange(desc(abs(difference)))

comparison_df
```

```{r feature-diff-plot}
comparison_df %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(feature, abs(difference)), y = difference,
             fill = difference > 0)) +
  geom_col(alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c("#d7191c", "#2c7bb6"),
                    labels = c("Below normal mean", "Above normal mean")) +
  labs(
    title = "Top 15 Features: Mean Difference (Flagged vs. Normal)",
    x     = NULL,
    y     = "Mean Difference",
    fill  = NULL
  ) +
  theme_minimal(base_size = 13)
```

---

# 10. Export Submission CSV

```{r export}
submission <- top100 %>%
  select(Rank, Transaction_ID, Amount)

write_csv(submission, "LOFRANKED>.csv")

cat("Submission saved: LOFRANKED>.csv\n")
cat("Rows:", nrow(submission), "\n")
head(submission, 10)
```

---

# 11. Summary

```{r summary-table}
tibble(
  Step        = c("Dataset size", "Features used", "LOF k values tested",
                  "Final model", "Transactions flagged",
                  "LOF score (k=20) range of top 100"),
  Detail      = c(
    paste(nrow(df_raw), "transactions"),
    "X01–X28 + Amount (scaled)",
    "k = 10, 20, 50",
    "Ensemble rank average of k=10, 20, 50",
    "100",
    paste0(round(min(top100$lof_k20), 2), " – ", round(max(top100$lof_k20), 2))
  )
) %>%
  knitr::kable(col.names = c("Item", "Detail"), align = "ll")
```

> **Next steps:** Replace `team_<your-team-name>.csv` with your actual team name before submitting. Check the `/submission/` example in the repo to confirm column formatting matches.
